Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2025-03-22 manuscript.Rmd
abstract shortening and csl modifications

Investigations into the environmental and anthropogenic factors that shape coral reef communities have been historically limited both in depth range and taxonomic scope. Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m of their depth ranges. Here, we examine the variability patterns of brachyuran crab assemblages recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth. We examine the effects of a suite of environmental, ecological, and anthropogenic factors (latitude, longitude, site depth, deployment years, chlorophyll-A, sea surface temperature, island slope, coral cover, potential larval immigration, cumulative human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblage structure was significantly different between the inhabited Main and relatively pristine Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, potential larval immigration, and depth. Despite this spatial and environmental variation, community differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.

>>>>

Investigations into the factors that shape coral reef communities have been historically limited in depth and taxonomic scope. Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m. Here, we examine the variability of crab assemblages from Autonomous Reef Monitoring Structures across Hawai&#x02BB;i and ~90 m depth. We examine the effects of environmental, ecological, and anthropogenic factors (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a mesophotic depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth. Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.
